Reanalysis Step 5: Outcome Analysis with Full Four-Treatment Interaction

### 2025-09-02 reanalysis-2-multiple_imputation.qmd
fix fix fix fix

# First, create person-level DR severity (maximum observed per patient)
df_person_dr <- df_for_imputation |>
  group_by(e_mrn_deidentified) |>
  summarise(
    # Take the maximum observed person_dr for each patient
    # This ensures consistency within patients
    person_dr_max = max(person_dr, na.rm = TRUE),
    # If all values are NA for a patient, max returns -Inf, so handle that
    person_dr_max = ifelse(is.infinite(person_dr_max), NA_real_, person_dr_max)
  ) |>
  ungroup()

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = row_number(),
    # Create numeric cluster ID for 2l.pan (required for multilevel imputation)
    cluster_id = as.numeric(factor(e_mrn_deidentified))
  ) |>
  # Join the person-level DR severity
  left_join(df_person_dr, by = "e_mrn_deidentified") |>
  dplyr::select(row_id, cluster_id, dplyr::all_of(imputation_vars), person_dr_max) |>
  mutate(
    # Convert person_dr_max to ordered factor for polr imputation method
    person_dr_max = factor(person_dr_max, levels = c(0, 1, 2), 
                          labels = c("No_DR", "NPDR", "PDR"),
                          ordered = TRUE),
    # Convert original person_dr to factor (keep for comparison but won't impute)
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    cohort_id = factor(cohort_id),  # Ensure cohort_id is a factor
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat),

@@

log_message(paste0("Created person_dr_max: ", 
                  sum(!is.na(df_mice$person_dr_max)), " non-missing, ",
                  sum(is.na(df_mice$person_dr_max)), " missing"))
```

@@

# Use proportional odds model for ordinal person_dr_max (0/1/2)
# person_dr_max is the person-level DR severity (consistent within patients)
custom_methods["person_dr_max"] <- "polr"  # Ordinal logistic regression
# Don't impute the original eye-level person_dr - it's replaced by person_dr_max
custom_methods["person_dr"] <- ""  # Don't impute eye-level DR

@@

# Handle person_dr and person_dr_max
# person_dr (eye-level) should not be used as predictor or predicted
pred_matrix["person_dr", ] <- 0  # Not predicted
pred_matrix[, "person_dr"] <- 0  # Not a predictor

# person_dr_max (person-level) should be used as predictor but only for appropriate variables
# It can predict outcomes but not baseline variables
pred_matrix[, "person_dr_max"] <- 1  # Use as predictor for most variables
pred_matrix["person_dr_max", "person_dr_max"] <- 0  # Not its own predictor
pred_matrix["baseline_VA_logMAR", "person_dr_max"] <- 0  # Don't use to predict baseline VA
pred_matrix["row_id", "person_dr_max"] <- 0  # Don't use for IDs
pred_matrix["e_mrn_deidentified", "person_dr_max"] <- 0  # Don't use for IDs
pred_matrix["cohort_id", "person_dr_max"] <- 0  # Don't use for cohort_id

@@

# Log person_dr_max imputation status
if ("person_dr_max" %in% names(mice_results$imp)) {
  n_imputed_dr <- sum(mice_results$nmis["person_dr_max"])
  log_message(paste0("person_dr_max imputed for ", n_imputed_dr, " observations"))
}

@@

  # CRITICAL: Copy person_dr_max (imputed) to person_dr for create_dr_severity_variables
  # person_dr_max is the person-level DR that was imputed
  # person_dr needs to be set for each eye from person_dr_max
  if ("person_dr_max" %in% names(imp_data)) {
    # Convert from factor if needed
    if (is.factor(imp_data$person_dr_max)) {
      # Convert factor levels (No_DR=1, NPDR=2, PDR=3) to numeric (0, 1, 2)
      imp_data$person_dr <- as.numeric(imp_data$person_dr_max) - 1
    } else {
      imp_data$person_dr <- imp_data$person_dr_max
    }
  }

@@

# FINAL STEP: Enforce patient-level variable consistency across all datasets
# person_dr, CCI, and DCSI should be consistent within patients
imputed_datasets <- lapply(imputed_datasets, function(imp_data) {
  # Calculate consistent values per patient (using max for all)
  patient_level_vars <- imp_data |>
    group_by(e_mrn_deidentified) |>
    summarise(
      # Handle case where all values are NA (max returns -Inf)
      person_dr_consistent = ifelse(all(is.na(person_dr)), 
                                   NA_real_, 
                                   max(person_dr, na.rm = TRUE)),
      CCI_consistent = ifelse(all(is.na(CCI)), 
                             NA_real_, 
                             max(CCI, na.rm = TRUE)),
      DCSI_consistent = ifelse(all(is.na(DCSI)), 
                              NA_real_, 
                              max(DCSI, na.rm = TRUE)),
      .groups = 'drop'
    )

  # Replace with consistent values
  imp_data <- imp_data |>
    left_join(patient_level_vars, by = "e_mrn_deidentified") |>
    mutate(
      person_dr = person_dr_consistent,
      CCI = CCI_consistent,
      DCSI = DCSI_consistent
    ) |>
    select(-person_dr_consistent, -CCI_consistent, -DCSI_consistent)

@@

log_message("Enforced patient-level consistency for person_dr, CCI, and DCSI by taking max value per patient")

>>>>

# Prepare data for MICE - NO person_dr_max, use eye-level person_dr
df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = row_number(),
    # Create numeric cluster ID for 2l.pan (required for multilevel imputation)
    cluster_id = as.numeric(factor(e_mrn_deidentified))
  ) |>
  dplyr::select(row_id, cluster_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    # person_dr is eye-level DR severity (0=No_DR, 1=NPDR, 2=PDR)
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    cohort_id = factor(cohort_id),  # Ensure cohort_id is a factor
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat),

@@

log_message(paste0("person_dr (eye-level): ", 
                  sum(!is.na(df_mice$person_dr)), " non-missing, ",
                  sum(is.na(df_mice$person_dr)), " missing"))
```

@@

# Use proportional odds model for ordinal person_dr (0/1/2)
# person_dr is the eye-level DR severity
custom_methods["person_dr"] <- "polr"  # Ordinal logistic regression

@@

# Handle person_dr (eye-level DR severity)
# person_dr should be used as predictor and can be predicted
pred_matrix[, "person_dr"] <- 1  # Use as predictor for most variables
pred_matrix["person_dr", "person_dr"] <- 0  # Not its own predictor
pred_matrix["baseline_VA_logMAR", "person_dr"] <- 0  # Don't use to predict baseline VA
pred_matrix["row_id", "person_dr"] <- 0  # Don't use for IDs
pred_matrix["e_mrn_deidentified", "person_dr"] <- 0  # Don't use for IDs
pred_matrix["cohort_id", "person_dr"] <- 0  # Don't use for cohort_id

@@

# Log person_dr imputation status
if ("person_dr" %in% names(mice_results$imp)) {
  n_imputed_dr <- sum(mice_results$nmis["person_dr"])
  log_message(paste0("person_dr imputed for ", n_imputed_dr, " observations"))
}

@@

  # Convert person_dr from factor back to numeric if needed
  if (is.factor(imp_data$person_dr)) {
    # Convert factor levels (No_DR=1, NPDR=2, PDR=3) to numeric (0, 1, 2)
    imp_data$person_dr <- as.numeric(imp_data$person_dr) - 1
  }

@@

# FINAL STEP: Enforce patient-level variable consistency across all datasets
# person_dr, CCI, and DCSI should be consistent within patients
imputed_datasets <- lapply(imputed_datasets, function(imp_data) {
  # Calculate consistent values per patient (using max for all)
  patient_level_vars <- imp_data |>
    group_by(e_mrn_deidentified) |>
    summarise(
      # For CCI and DCSI, take minimum (better value) per patient
      # Lower scores indicate better health
      CCI_consistent = ifelse(all(is.na(CCI)), 
                             NA_real_, 
                             min(CCI, na.rm = TRUE)),
      DCSI_consistent = ifelse(all(is.na(DCSI)), 
                              NA_real_, 
                              min(DCSI, na.rm = TRUE)),
      # Note: person_dr is now eye-level, so we don't enforce consistency
      .groups = 'drop'
    )

  # Replace with consistent values (CCI and DCSI only, not person_dr)
  imp_data <- imp_data |>
    left_join(patient_level_vars, by = "e_mrn_deidentified") |>
    mutate(
      # Only replace CCI and DCSI with consistent (minimum) values
      CCI = coalesce(CCI_consistent, CCI),
      DCSI = coalesce(DCSI_consistent, DCSI)
      # person_dr remains eye-level, not enforced to be consistent
    ) |>
    select(-CCI_consistent, -DCSI_consistent)

@@

log_message("Enforced patient-level consistency for CCI and DCSI by taking minimum (better) value per patient")

### 2025-09-02 reanalysis-5-outcome_analysis-full_interaction.qmd
fix fix fix fix

---
title: "Reanalysis Step 5: Outcome Analysis with Three-Way Interaction"
subtitle: "IPTW-weighted analysis of lapse × DR severity × treatment interactions"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

# Initialize logging
log_file <- init_log("outcome_analysis_interaction")
```

@@

```{r create-variables}
# Create consistent analysis variables across all imputed datasets
for (i in 1:n_imputations) {
  imputed_datasets[[i]] <- imputed_datasets[[i]] |>
    mutate(
      # Binary treatment indicator (any treatment vs none)
      any_treatment = as.numeric(anti_VEGF == 1 | PRP_flag == 1 | 
                                 other_inject == 1 | focal_laser_flag == 1),
      any_treatment = factor(any_treatment, levels = c(0, 1), 
                            labels = c("No_Treatment", "Any_Treatment")),

@@

@@

```{r define-formula}
# Get the interaction formulas from functions.R
interaction_formulas <- get_analysis_formulas_reanalysis_interaction()

# Use the full formula which includes three-way interaction
interaction_formula <- interaction_formulas$full

@@

**Formula:**
```{r show-formula}
print(interaction_formula)
```

@@

```{r fit-models}
# Check if models are already cached
models_cache_file <- file.path(reanalysis_data_dir, "fitted_models_interaction.rds")

if (file.exists(models_cache_file) && !force_refit_models) {
  # Load cached models
  cached_models <- readRDS(models_cache_file)
  models <- cached_models$models
  model_summaries <- cached_models$model_summaries
  model_progress <- cached_models$model_progress
  models_loaded_from_cache <- TRUE
  cache_location <- models_cache_file
  log_message("Models loaded from cache: ", cache_location)
} else {
  # Fit new models
  models_loaded_from_cache <- FALSE
  models <- list()
  model_summaries <- list()

  model_progress <- data.frame(
    Imputation = integer(),
    Converged = logical(),
    ThreeWay_Terms = integer()
  )

  for (i in 1:n_imputations) {

@@

  model <- svyglm(
    formula = interaction_formula,
    design = design_weighted,
    family = quasibinomial()
  )

  models[[i]] <- model

  coef_summary <- summary(model)$coefficients

  # Check if three-way interaction terms exist
  threeway_terms <- grep("ever_lapse_binary.*:dr_severity.*:any_treatment", 
                         rownames(coef_summary), value = TRUE)

  # Track progress
  model_progress <- rbind(model_progress, data.frame(
    Imputation = i,
    Converged = model$converged,
    ThreeWay_Terms = length(threeway_terms)
  ))

  model_summaries[[i]] <- coef_summary
  }

  # Save to cache
  cached_models <- list(
    models = models,
    model_summaries = model_summaries,
    model_progress = model_progress,
    formula_used = interaction_formula,
    n_imputations = n_imputations,
    timestamp = Sys.time()
  )

>>>>

---
title: "Reanalysis Step 5: Outcome Analysis with Full Four-Treatment Interaction"
subtitle: "IPTW-weighted analysis of lapse × DR severity × 4-level treatment interactions"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

# Initialize logging
log_file <- init_log("outcome_analysis_full_interaction")

# Configuration for stability checking
MAX_SE_THRESHOLD <- 10  # Maximum acceptable standard error
MIN_N_PER_CELL <- 5    # Minimum observations per interaction cell
TEST_SINGLE_IMPUTATION <- TRUE  # Test on single imputation first
```

@@

```{r create-variables}
# Test on single imputation first if configured
if (TEST_SINGLE_IMPUTATION) {
  n_test <- 1
  log_message("TESTING MODE: Will run on single imputation first")
} else {
  n_test <- n_imputations
}

# Create consistent analysis variables across all imputed datasets
for (i in 1:n_imputations) {
  imputed_datasets[[i]] <- imputed_datasets[[i]] |>
    mutate(
      # Create 4-level treatment factor (hierarchical priority)
      # Priority: anti_VEGF > PRP > focal_laser > other_inject > none
      treatment_type = case_when(
        anti_VEGF == 1 ~ "anti_VEGF",
        PRP_flag == 1 ~ "PRP",
        focal_laser_flag == 1 ~ "focal_laser",
        other_inject == 1 ~ "other_inject",
        TRUE ~ "none"
      ),
      treatment_type = factor(treatment_type, 
                              levels = c("none", "other_inject", "focal_laser", 
                                       "PRP", "anti_VEGF")),

      # Keep binary for comparison
      any_treatment = as.numeric(anti_VEGF == 1 | PRP_flag == 1 | 
                                 other_inject == 1 | focal_laser_flag == 1),
      any_treatment = factor(any_treatment, levels = c(0, 1), 
                            labels = c("No_Treatment", "Any_Treatment")),

@@

# Check treatment distribution
treatment_dist <- table(imputed_datasets[[1]]$treatment_type)
log_message("Treatment type distribution in first imputation:")
for (tx in names(treatment_dist)) {
  log_message(paste0("  ", tx, ": ", treatment_dist[tx]))
}
```

## Check Cell Sizes for Full Interaction

```{r check-cells}
# Check cell sizes for the three-way interaction
first_data <- imputed_datasets[[1]]
cell_counts <- first_data |>
  count(ever_lapse_binary, dr_severity, treatment_type) |>
  arrange(n)

# Identify problematic cells
small_cells <- cell_counts |> filter(n < MIN_N_PER_CELL)

if (nrow(small_cells) > 0) {
  log_message(paste0("WARNING: ", nrow(small_cells), 
                    " cells have fewer than ", MIN_N_PER_CELL, " observations"))

  print(kable(small_cells, 
              caption = paste("Cells with < ", MIN_N_PER_CELL, " observations")))

  # Decide whether to proceed
  if (nrow(small_cells) > length(unique(first_data$treatment_type))) {
    warning("Many small cells detected. Model may be unstable.")
    # Could stop here or collapse categories
  }
}

# Display full interaction table
interaction_table <- cell_counts |>
  pivot_wider(
    names_from = treatment_type,
    values_from = n,
    values_fill = 0
  ) |>
  arrange(ever_lapse_binary, dr_severity)

kable(interaction_table,
      caption = "Cell sizes for Lapse × DR Severity × Treatment Type (4 levels)")
```

@@

```{r define-formula}
# Define the full interaction formula with 4-level treatment
# This creates a much more complex model than the binary treatment version
full_interaction_formula <- as.formula(
  "outcome_va_vi_binary ~ 
    ever_lapse_binary * dr_severity * treatment_type +  # Full 3-way with 4-level treatment
    baseline_VA_logMAR +
    gender_cat + race_ethnic_cat + insurance_cat + age_cat +
    CCI + DCSI +
    glaucoma_bef_hitplus_cat + otherretina_bef_hitplus_cat + 
    catsurg_before_hitplus_cat"
)
```

@@

**Formula:**
```{r show-formula}
print(full_interaction_formula)

# Calculate expected number of terms
# Main effects: 1 (lapse) + 2 (DR) + 4 (treatment) = 7
# Two-way: 1*2 + 1*4 + 2*4 = 2 + 4 + 8 = 14  
# Three-way: 1*2*4 = 8
# Total interaction terms = 7 + 14 + 8 = 29 (plus covariates)
expected_terms <- 29
log_message(paste0("Expected interaction terms: ", expected_terms))
```

@@

```{r fit-models}
# Check if models are already cached
models_cache_file <- file.path(reanalysis_data_dir, "fitted_models_full_interaction.rds")

# Determine how many models to fit
n_to_fit <- if (TEST_SINGLE_IMPUTATION) 1 else n_imputations

if (file.exists(models_cache_file) && !force_refit_models && !TEST_SINGLE_IMPUTATION) {
  # Load cached models
  cached_models <- readRDS(models_cache_file)
  models <- cached_models$models
  model_summaries <- cached_models$model_summaries
  model_progress <- cached_models$model_progress
  models_loaded_from_cache <- TRUE
  cache_location <- models_cache_file
  log_message("Models loaded from cache: ", cache_location)
} else {
  # Fit new models
  models_loaded_from_cache <- FALSE
  models <- list()
  model_summaries <- list()

  model_progress <- data.frame(
    Imputation = integer(),
    Converged = logical(),
    ThreeWay_Terms = integer(),
    Max_SE = numeric(),
    Warning = character()
  )

  for (i in 1:n_to_fit) {
    log_message(paste0("Fitting model for imputation ", i, " of ", n_to_fit))

@@

    # Fit with error handling
    fit_success <- FALSE
    warning_msg <- ""

    tryCatch({
      model <- svyglm(
        formula = full_interaction_formula,
        design = design_weighted,
        family = quasibinomial()
      )

      fit_success <- TRUE

      # Check model stability
      coef_summary <- summary(model)$coefficients
      max_se <- max(coef_summary[, "Std. Error"], na.rm = TRUE)

      if (max_se > MAX_SE_THRESHOLD) {
        warning_msg <- paste0("High SE: ", round(max_se, 2))
        problem_terms <- rownames(coef_summary)[coef_summary[, "Std. Error"] > MAX_SE_THRESHOLD]
        log_message(paste0("WARNING: Imputation ", i, " has extreme SEs in: ", 
                          paste(problem_terms, collapse = ", ")))
      }

    }, error = function(e) {
      warning_msg <- paste0("Fit failed: ", e$message)
      log_message(paste0("ERROR: Imputation ", i, " - ", warning_msg))
    })

    if (fit_success) {
      models[[i]] <- model

      # Check if three-way interaction terms exist (with 4-level treatment)
      threeway_terms <- grep("ever_lapse_binary.*:dr_severity.*:treatment_type", 
                           rownames(coef_summary), value = TRUE)

      # Track progress
      model_progress <- rbind(model_progress, data.frame(
        Imputation = i,
        Converged = model$converged,
        ThreeWay_Terms = length(threeway_terms),
        Max_SE = max_se,
        Warning = warning_msg,
        stringsAsFactors = FALSE
      ))

      model_summaries[[i]] <- coef_summary

      # Stop if testing and issues found
      if (TEST_SINGLE_IMPUTATION && i == 1 && max_se > MAX_SE_THRESHOLD) {
        print(summary(model))
        stop(paste0("Model instability detected. Max SE = ", round(max_se, 2), 
                   ". Consider simplifying model or collapsing treatment categories."))
      }
    }
  }

  # Save to cache
  cached_models <- list(
    models = models,
    model_summaries = model_summaries,
    model_progress = model_progress,
    formula_used = full_interaction_formula,
    n_imputations = n_imputations,
    timestamp = Sys.time()
  )
